MolProbity's Ultimate Rotamer-Library Distributions for Model Validation

### 2016-02-16 response_2_reviwers.txt

The following details concerns brought up by the reviewers and our response.

Reviewer 1:
- Could you devise a figure or table that more directly or comprehensively compares the older (penultimate) rotamer library to the new (ultimate) you describe here

In the supplement we added a table that compares the outlier count in the unfiltered dataset when evaluated using the Top500 vs. the Top8000 contours (see table 2 in supplement). We have also added a figure comparing the contour of the remaining 2 dimensional residue types not already in figure 2 (see figure 1 in the supplement). Along with this we added references to Figure S1 in 4.2 and the caption of Figure 2 (See red text).

@@

The following citation to SideAide was added in the introduction where talking about where rotamer libraries are used (brackets enclosing this citation are in red):

R. P. Joosten, K. Joosten, S. X. Cohen, G. Vriend, and A. Perrakis. Automatic rebuild- ing and optimization of crystallographic structures in the protein data bank. Bioinfor- matics, 27(24):3392–3398, oct 2011.

- A reference to RSCC is missing (page 6)

    There seems to be no definitive reference that claims development of the RSCC, as we concluded after an extensive search of the literature and the relevant web sites, and consulting the other Phenix PIs.  Jones et al (2004) Acta Cryst A47:110-9 introduces the real-space residual (RSR), but only mentions possible use of a correlation function in passing.  The Electron Density Server presents both RSR and RSCC values for each structure, and in Kleywegt et al (2004) Acta Cryst D60:2240-9 they describe the calculation as being done in MAPMAN (the paper for which does not mention RSCC) and give more additional details than anywhere else,  but not quite enough to reproduce it.  That's why we gave the equation and parameters in our paper.  If you know of the right reference, please let us know.
    Here, we have changed the detailed description in 3.2 methods (p. 8) to include:
"… a local real-space correlation coefficient (RSCC) metric [Kleywegt 2004]" (brackets enclosing this citation are in red)
"… how the RSCC was calculated, as implemented standardly in Phenix." (see red text)

### 2016-02-16 methods.tex
added RSCC reference

\subsection{Residue-level Dataset Filters}
\label{sec:resfilters}
Although the chain-level filters select for structures with high overall quality, local quality varies considerably within any model. In this context, \emph{quality} refers to canonical macromolecular model validation criteria (sterics, geometry, and conformation) as well as the fit to electron density. An important aspect of this research is to include only amino-acid sidechains with sufficiently clear electron density to justify their given conformation. Our previous rotamer library was based on the Top500 dataset where only two residue-level filters were used: no clashes and no B factors $\ge$ 40 \AA\textsuperscript{2}. The B factor filter was then (in 2003) our best available proxy for electron-density quality; however, residues with dubious sidechain density still remained. To remedy this, the current dataset was also filtered by a local real-space correlation coefficient (RSCC) metric and by local map value.

@@

The local RSCC looks at the correlation between the $\sigma_{A}$-weighted \EDmap{} and the F$_{c}$ (calculated from model) electron density maps for a given local region, in this case, around each individual atom. Equation \ref{rscceq} shows how the RSCC was calculated. $o$ and $c$ are $\sigma$ values at grid points in the \EDmap{} and the F$_{c}$ maps, respectively, in a radius around the atom. The radius used is resolution dependent; a 1\AA{} radius is used for resolutions below 1\AA{} and a 1.5\AA{} radius is used for resolutions between 1\AA{} and 2\AA{}. $n$ is the number of grid points in the selected region and $\bar{o}$ and $\bar{c}$ are the local mean values for $o$ and $c$, respectively. The RSCC alone is not an adequate density-fit metric for our purposes. However, local RSCC (for shape), local \EDmap{} $\sigma$ value at the atom coordinate (for height), and B factor (for spread) together create a satisfactory fit-to-density metric. We used \textit{phenix.real\_space\_correlation} in the \textit{PHENIX} software package \citep{Adams:2010fk} to calculate the RSCC and \EDmap{} $\sigma$ values at each atom for all structures in the Top8000. Per-residue filter values were then assigned by taking the worst atom B (greatest value), worst atom RSCC (least value), and worst atom \EDmap{} $\sigma$ value (least value) in each residue.

>>>>

\subsection{Residue-level Dataset Filters}
\label{sec:resfilters}
Although the chain-level filters select for structures with high overall quality, local quality varies considerably within any model. In this context, \emph{quality} refers to canonical macromolecular model validation criteria (sterics, geometry, and conformation) as well as the fit to electron density. An important aspect of this research is to include only amino-acid sidechains with sufficiently clear electron density to justify their given conformation. Our previous rotamer library was based on the Top500 dataset where only two residue-level filters were used: no clashes and no B factors $\ge$ 40 \AA\textsuperscript{2}. The B factor filter was then (in 2003) our best available proxy for electron-density quality; however, residues with dubious sidechain density still remained. To remedy this, the current dataset was also filtered by a local real-space correlation coefficient (RSCC) metric \textcolor{red}{\cite{Kleywegt:ba5061}} and by local map value.

@@

The local RSCC looks at the correlation between the $\sigma_{A}$-weighted \EDmap{} and the F$_{c}$ (calculated from model) electron density maps for a given local region, in this case, around each individual atom. Equation \ref{rscceq} shows how the RSCC was calculated\textcolor{red}{, as implemented in \textit{PHENIX}}. $o$ and $c$ are $\sigma$ values at grid points in the \EDmap{} and the F$_{c}$ maps, respectively, in a radius around the atom. The radius used is resolution dependent; a 1\AA{} radius is used for resolutions below 1\AA{} and a 1.5\AA{} radius is used for resolutions between 1\AA{} and 2\AA{}. $n$ is the number of grid points in the selected region and $\bar{o}$ and $\bar{c}$ are the local mean values for $o$ and $c$, respectively. The RSCC alone is not an adequate density-fit metric for our purposes. However, local RSCC (for shape), local \EDmap{} $\sigma$ value at the atom coordinate (for height), and B factor (for spread) together create a satisfactory fit-to-density metric. We used \textit{phenix.real\_space\_correlation} in the \textit{PHENIX} software package \citep{Adams:2010fk} to calculate the RSCC and \EDmap{} $\sigma$ values at each atom for all structures in the Top8000. Per-residue filter values were then assigned by taking the worst atom B (greatest value), worst atom RSCC (least value), and worst atom \EDmap{} $\sigma$ value (least value) in each residue.
